---
title: "Access to Public Transit in Boston"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
---

```{r setup, include=FALSE}
library(flexdashboard)
```

```{r, include=FALSE}
# Setting working directory 
setwd("~/Desktop/FinalProj")

# Including libraries
library(sf)
library(dplyr)
library(tmap)
library(ggplot2)
library(tidycensus)
library(raster)
library(MASS)
library(leaflet)
library(ggmap)
library(ggspatial)
library(ggsn)
library(plotly)

# enabling tmap interactive mode
tmap_mode("view")

## API keys not publicly exposed ##
```

```{r, include=FALSE}
# -------Base maps---------

# Boston neighborhoods
neighborhoods <- st_read("Census2020_BG_Neighborhoods.shp")

# Neighborhood geometries in NAD83
neighborhoods_N83 <- st_transform(neighborhoods, st_crs(4269))

# Mass municipalities polygons
mass_poly <- st_read("TOWNSSURVEY_POLY.shp")

# Filter by Boston 
mass_poly_bos <- mass_poly %>% filter(TOWN==c("BOSTON"))

# Crop view 
mass_poly_bos_cropped <- st_crop(mass_poly_bos, xmin = 220391.7, xmax = 241871.9,
                                    ymin = 894241, ymax = 909753.4)

# Cropped version in NAD83
mass_poly_bos_cropped_N83 <- st_transform(mass_poly_bos_cropped, crs=st_crs(4269))
```

```{r}
# Stamen

us <- c(left = -71.18, bottom = 42.3, right = -71.0, top = 42.39)
boston_base <- get_stamenmap(us, zoom=14, maptype = 'terrain-background', source = 'stamen') 

```

```{r, include=FALSE}
# -------MBTA data---------

# Subway station locations
t_sub_nodes <- st_read("MBTA_NODE.shp")
# Subway station lines
t_sub_lines <- st_read("MBTA_ARC.shp")

# Bus stops
t_bus_stops <- st_read("MBTABUSSTOPS_PT.shp")

# Subway lines (note: this does not include the entire ARC namespace)
t_sub_lines_GREEN <- t_sub_lines %>% filter(t_sub_lines$LINE=="GREEN")       
t_sub_lines_ORANGE <- t_sub_lines %>% filter(t_sub_lines$LINE=="ORANGE") 
t_sub_lines_RED <- t_sub_lines %>% filter(t_sub_lines$LINE=="RED") 
t_sub_lines_SILVER <- t_sub_lines %>% filter(t_sub_lines$LINE=="SILVER") 
t_sub_lines_BLUE <- t_sub_lines %>% filter(t_sub_lines$LINE=="BLUE") 

```

```{r, include=FALSE}
## NAD83 / Massachusetts Mainland

# Extracting shapefile geometries into a separate data frame
t_sub_nodes_points <- as.data.frame(st_coordinates(st_geometry(t_sub_nodes)))
t_bus_stops_points <- as.data.frame(st_coordinates(st_geometry(t_bus_stops)))

t_sub_bus_points <- rbind(t_sub_nodes_points, t_bus_stops_points)

## WG84

# Conversion
t_sub_nodes_W84 <- st_transform(t_sub_nodes, st_crs(4326))
t_bus_stops_W84 <- st_transform(t_bus_stops, st_crs(4326))

# Extracting shapefile geometries into a separate data frame
t_sub_nodes_W84_points <- as.data.frame(st_coordinates(st_geometry(t_sub_nodes_W84)))
t_bus_stops_W84_points <- as.data.frame(st_coordinates(st_geometry(t_bus_stops_W84)))

t_sub_bus_W84_points <- rbind(t_sub_nodes_W84_points, t_bus_stops_W84_points)

## NAD83

# Conversion
t_sub_nodes_N83 <- st_transform(t_sub_nodes, st_crs(4269))
t_bus_stops_N83 <- st_transform(t_bus_stops, st_crs(4269))

# Extracting shapefile geometries into a separate data frame
t_sub_nodes_N83_points <- as.data.frame(st_coordinates(st_geometry(t_sub_nodes_N83)))
t_bus_stops_N83_points <- as.data.frame(st_coordinates(st_geometry(t_bus_stops_N83)))

t_sub_bus_N83_points <- rbind(t_sub_nodes_N83_points, t_bus_stops_N83_points)
```

```{r, include=FALSE}
## Census data
census_tract_data <- get_acs(geography = "tract",
                      variables = c(total_pop = "B01003_001"),
                      state = "MA",
                      county = "Suffolk",
                      geometry = TRUE,
                      year = 2019)

census_tract_data$"Population Estimate" <- census_tract_data$estimate
class(census_tract_data$"Population Estimate")

```

```{r, include=FALSE}
# -------Kernel Density Estimate (Rapid Transit)---------
# with assistance from "Making Sense Remotely" 

# 2D Kernel Density Estimate 
kernel_density <- kde2d(t_sub_nodes_points$X, t_sub_nodes_points$Y, n = 100)
contour(kernel_density)

# Kernel density estimate raster data
kde_raster = raster(kernel_density)
raster_df = as.data.frame(kde_raster, xy=T)

# Highest density profile (find maximum density within each group)
highest_dens_prof <- aggregate(raster_df$layer, by=list(raster_df$x), FUN = max)

# Gets the max density values from original data frame
highest_dens_prof <- raster_df[match(highest_dens_prof$x, raster_df$layer), ]

# Plot primitive
#lines(highest_dens_prof$x, highest_dens_prof$y, col = "orange", lwd=2)
```

```{r, include=FALSE}
# -------NAD83 Kernel Density Estimate (Rapid Transit)---------

# 2D Kernel Density Estimate
kernel_density_83 <- kde2d(t_sub_nodes_N83_points$X, t_sub_nodes_N83_points$Y, n = 100)
contour(kernel_density_83)

# Kernel density estimate raster data
kde_raster_83 = raster(kernel_density_83)
raster_df_83 = as.data.frame(kde_raster_83, xy=T)

# Highest density profile (find maximum density within each group)
highest_dens_prof_83 <- aggregate(raster_df_83$layer, by=list(raster_df_83$x), FUN = max)

# Gets the max density values from original data frame
highest_dens_prof_83 <- raster_df_83[match(highest_dens_prof_83$x, raster_df_83$layer), ]

# Plot primitive
#lines(highest_dens_prof_83$x, highest_dens_prof_83$y, col = "orange", lwd=2)
```

```{r, include=FALSE}
## NAD83 length unit is meter
## 0.25 vmiles is 804.672 meters
## from https://safety.fhwa.dot.gov/ped_bike/ped_transit/ped_transguide/ch4.cfm#:~:text=A.,when%20accessing%20heavy%20rail%20services.

t_sub_nodes_N83_buffer_0.25 <- st_buffer(t_sub_nodes_N83, 402.336)
```


```{r, include=FALSE}
stations_per_neighborhood <- neighborhoods_N83 %>% 
  mutate("Subway_Station_Count" = lengths(st_intersects(., t_sub_nodes_N83)))

stations_per_neighborhood <- cbind(label = paste("Number of stations in ",
                    stations_per_neighborhood$BlockGr202, 
                    ": ", 
                    stations_per_neighborhood$Subway_Station_Count,
                    sep = ""), stations_per_neighborhood)
stations_per_neighborhood %>% filter(stations_per_neighborhood$Subway_Station_Count==0)
```

Rapid Density {data-icon="fa-signal"}
===================================== 

### Station Density Map (MBTA Rapid Transit) 
    
```{r}
# IN NAD83 MASS MAINLAND / NO NEIGHBORHOODS

mass_poly_bos_cropped <- sf::st_cast(mass_poly_bos_cropped, "MULTIPOLYGON")

T_dens_no_neighborhoods <- ggplot() +  
  geom_sf(data=mass_poly_bos_cropped) +
  stat_density_2d(data=t_sub_nodes_points, geom="polygon", aes(x=X, y=Y, fill = after_stat(1000000000*level))) +
  scale_fill_viridis_c(alpha=0.3)+
  xlim(226392, 241841) +
  ylim(895241, 905753) +
  geom_line(aes(x = highest_dens_prof$x, y = highest_dens_prof$y, color = "HDP"), linetype="dotted", 
            size=1, labels = "Highest density profile") + 
  geom_point(aes(x=236552, y=900671), colour="black") +
  labs(x = "", y = "", colour = "Highest desnity profile", fill = "Station density values \n(NAD83 / Massachusetts \nMainland)") +
  scale_color_manual(values=c("HDP"="#FFA000")) + 
  north(mass_poly_bos_cropped, location = "bottomleft", symbol = 3, anchor = c(x=225479, y=895641), scale=0.15) +
  theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5), 
        legend.title=element_text(size=9),
        plot.caption = element_text(hjust = 0),
        plot.subtitle = element_text(hjust = 0.5, size = 8, color = "darkgray")) + annotate(
   geom = "segment", x = 233400, y = 902200, xend = 236552, yend = 900671) + 
   annotate(geom = "text", x = 232892, y = 903000, label = "Downtown", size = 3) +
   annotate(geom = "text", x = 232892, y = 902500, label = "Boston", size = 3) +
   annotation_scale()
```

```{r}
ggplotly(T_dens_no_neighborhoods) %>% 
  layout(xaxis = list(autorange = TRUE),
         yaxis = list(autorange = TRUE))
```

Combined Density {data-icon="fa-signal"}
===================================== 

### Station Density Map (MBTA Buses and Rapid Transit)
    
```{r}
T_dens_BUS_no_neighborhoods <- ggplot() +  
  geom_sf(data=mass_poly_bos_cropped) +
  stat_density_2d(data=t_sub_bus_points, geom="polygon", aes(x=X, y=Y, fill = after_stat(1000000000*level))) +
  scale_fill_viridis_c(alpha=0.3, limits=c(5,16), na.value = NA) +
  xlim(226392, 241841) +
  ylim(895241, 905753) +
  labs(x = "", y = "", fill = "Station density values \n(NAD83 / Massachusetts \nMainland)") +
  north(mass_poly_bos_cropped, location = "bottomleft", symbol = 3, anchor = c(x=225479, y=895641), scale=0.15) +
  theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5),
        legend.title=element_text(size=9),
        plot.caption = element_text(hjust = 0),
        plot.subtitle = element_text(hjust = 0.5, size = 8, color = "darkgray")) +
  annotation_scale()
```

```{r}
ggplotly(T_dens_BUS_no_neighborhoods) %>% 
  layout(xaxis = list(autorange = TRUE),
         yaxis = list(autorange = TRUE))
```


HDP vs. Population {data-icon="fa-signal"}
=====================================     

### Station Density Profile vs. Tract Population (MBTA Rapid Transit)
    
```{r}
## NO SMOOTHING HDP vs. pop

T_dens_line_pop <- ggplot() +  
  geom_sf(data=census_tract_data, aes(fill = census_tract_data$"Population Estimate")) +
  scale_fill_viridis_c(alpha = 0.5) +
  geom_line(aes(x = highest_dens_prof_83$x, y = highest_dens_prof_83$y, color = "HDP"), linetype="dotted",
            size=1.4, labels = "Highest density profile") + 
  labs(x = "", y = "", colour = "Highest desnity profile", fill = "Tract Population") +
  scale_color_manual(values=c("HDP"="#FFA000")) + 
  theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5),
        plot.caption = element_text(hjust = 0, size = 6),
        plot.subtitle = element_text(hjust = 0.5, size = 8, color = "darkgray")) +
  north(census_tract_data, symbol = 3, location = "topleft", anchor=c(x = -71.2177, y = 42.439),scale=0.13) +
  xlim(-71.23, -70.95) +
  ylim(42.28, 42.45) + annotation_scale(location="br") +
   annotate(geom = "segment", x = -71.1, y = 42.375, xend = -71.0572, yend = 42.3557) + 
   annotate(geom = "text", x = -71.109, y = 42.3875, label = "Downtown", size = 3) +
   annotate(geom = "text", x = -71.109, y = 42.380, label = "Boston", size = 3) 
```

```{r}
ggplotly(T_dens_line_pop) %>% 
  layout(xaxis = list(autorange = TRUE),
         yaxis = list(autorange = TRUE))
```

Buffer Map {data-icon="fa-signal"}
===================================== 

### Rapid Transit Stations Within Walking Distance

```{r}
# Generating a map of subway stations
tm_shape(t_sub_nodes_N83_buffer_0.25) + tm_borders()

```

KDE Raster {data-icon="fa-signal"}
===================================== 

### Station Density Raster (MBTA Rapid Transit)

```{r}

raster_colors <- colorNumeric(c("#440154", "#414487", "#2a788e", "#22a884", "#7ad151", "#fde725"), values(kde_raster_83),
  na.color = "transparent")

m <- leaflet() %>% setView(lng = -71.0789, lat = 42.3501, zoom = 13)

m %>% addProviderTiles(providers$CartoDB.Positron) %>%
  addRasterImage(kde_raster_83, colors = raster_colors, opacity = 0.4) %>%
  addLegend(pal = raster_colors, values = values(kde_raster_83),
    title = "Station density \n(NAD83)")

```

Neighborhoods {data-icon="fa-signal"}
===================================== 

### Number of Subway Stations by Neighborhood


```{r}
tm_shape(stations_per_neighborhood) + tm_borders() + tm_fill("Subway_Station_Count", palette = "viridis", alpha=0.5)

```

